\forestclaw: Hybrid forest-of-octrees AMR for hyperbolic conservation laws

### 2013-03-18 docs/parco13/abstract.tex
Minor changes

\usepackage{amsmath,amssymb}
\usepackage{cite}
\usepackage{color}
\usepackage{graphicx}
\usepackage{url}
\usepackage{xspace}
\usepackage{graphicx}

@@

\begin{keyword}
adaptive mesh refinement,
hyperbolic conservation laws,
clawpack,
HPC,
manycore
\end{keyword}

@@

With the advent of many-core chips, such as GPUs and the MIC architecture, comes the opportunity to sustain unprecedented rates of floating point
operations at comparably high integration density and low cost.  These
architectures, however, require careful structuring of the data layout and
memory access patterns to exhaust their multithreading and vectorization
capabilities.

Consequently, it is not clear a priori how to accelerate PDE solvers
that use adaptive mesh refinement.
%, especially when working with
%unstructured meshes \comment{This sounds we are now going to talk
%  about unstructured meshes.}.
% CB: Good point, took it out.
Of course, it was realized early that it helps to aggregate degrees of
freedom (DOF) at the element level, as has been done with high-order
spectral element \cite{TufoFischer99}, low order continuous Galerkin via methods that accumulate many elements simultaneously \cite{knepleyterrel:2013},
%finite volume (element?)
or discontinuous Galerkin \cite{HesthavenWarburton02}
methods.  GPU implementations of the latter have been proposed recently
\cite{KlocknerWarburtonBridgeEtAl09, BursteddeGhattasGurnisEtAl10}.
The finite volume method has typically been implemented using a single degree
of freedom per cell on
% either
structured
\cite{ppm, clawpack}
or unstructured meshes
\cite{li:2010, he-hu:2008, do-om:2005, li-sh-sv-va:2007, openfoam},
where higher order methods have been constructed by widening the stencil, for instance in WENO methods
\cite{Shu:2009bi}.
% \comment{Are you looking for
%   references to parallel AMR codes?  Or more general AMR codes based
%   on the FVM?}  \todo{No, I was looking for unstructured FV
%   references.  Would you have any, and can we add the block-AMR
%   references later?} \comment{But most finite volume methods
%   (structured or unstructured) only have a single degree of freedom
%   per mesh cell (``element''?)  Only recently have people increased
%   the order by extending the stencil (without increase the DOFs per
%   mesh cell).}
% CB: Can you fix some of the above citations?

@@

To facilitate hardware acceleration for parallel dynamic AMR, we
build upon the forest-of-octrees paradigm because of its low overhead
and proven scalability \cite{BursteddeWilcoxGhattas11}.  This approach
identifies each octree leaf with a mesh element.  In our work, we go
beyond the traditional high-order element and define each element to
be a dense computational patch with $m^d$ DOFs.  In fact, this approach
resembles block-structured AMR \cite{be-ol:1984, be-co:1989,
% be-le:1991 (cited below)
ColellaGravesKeenEtAl07,
BerzinsLuitjensMengEtAl10} except that the
patches are not overlapping,
% (except in ghost cell regions),
% CB: we comment on this later in Section 4.
which enables us to capitalize on our
previous experience with scalable FE solvers for PDEs
\cite{BursteddeStadlerAlisicEtAl13}.  The \clawpack software
\cite{LeVeque97} provides a popular implementation of such a patch.
It has been designed to solve hyperbolic conservation laws
% on a uniform compute patch
and successfully used in the context of block-structured AMR
\cite{be-le:1991, amrclaw, Berger:2011du}.

@@

The starting point of our work is defined by the \pforest algorithms for
forest-of-octrees AMR on the one hand, and the \clawpack algorithms for the
numerical solution of hyperbolic conservation laws
% on uniformly gridded domains
on the other.  Both are specialized codes with the following characteristics:
% \comment{I am not sure about the Python dependency;  the latest version of
% \clawpack relies on
%Python to generate an input file, but the code itself is F77 (or maybe some
%F90). }
% CB: I made it optional.
% DAC: Good.
\begin{center}
\begin{tabular}{l|l|l}
& \multicolumn{1}{c|}{\pforest} & \multicolumn{1}{c}{\clawpack} \\
\hline
subject & hexahedral nonconforming mesh &  hyperbolic PDE on $[0, 1]^d$ \\
toplevel unit & forest of octrees & patch of $m^d$ FV cells \\
atomic unit & octree leaf & one DOF in each cell \\
parallelization & MPI & threads (\manyclaw variant) \\
memory access & distributed & shared on each MPI rank \\
data type & integers & floating point values \\
language & C & Fortran 77 \\
% CB: dropped dependencies.
% dependencies & none & Python (optional) \\
%\hline
\end{tabular}
\end{center}
Each leaf as the atomic unit of \pforest houses a toplevel unit of \clawpack.
The term cell is used to identify a single DOF within a \clawpack patch.  The
proposed 1:1 correspondence between a leaf and a patch thus combines two
previously disjoint models in a modular way:
\begin{enumerate}
\item We permit the reuse of existing, verified, and performant codes.
% if the integration is done well
\item We preserve the separation between the mesh on one hand and the
discretization and solvers on the other.
\item The AMR metadata (\pforest:
under 1k bytes per octree, 8 bytes per MPI rank,
24 bytes per leaf. \forestclaw: $84 + 28d$ bytes per patch)
% \comment{?? : in 3d: 5 ints,  9 floats; in 2d: 4 ints, 6 floats}
% CB: thanks, I have added it.
is insignificant compared to
the numerical data ($m^d$ floating point values per patch).
\item The resulting parallel programming model is a hybrid (often referred
to as MPI+X).  Only rank-local leaves/patches are stored and computed on.
\end{enumerate}

@@

\begin{figure}
\begin{center}
\includegraphics[width=.4\columnwidth]{../figs_misc/forest_stacked}
\hspace{.05\columnwidth}
\includegraphics[width=.4\columnwidth]{../figs_misc/cubed3}
\end{center}
\caption{Left: Forest of two quadtrees, partitioned among three MPI processes.
Each quadtree has its own coordinate orientation.  The situation in 3D
(octrees) is analogous.  Right: The leaves in a forest of six quadtrees that
serves as the computational domain for the cubed sphere.  An adhoc refinement
pattern has been 2:1 balanced in terms of neighbor sizes and partitioned
between five MPI processes (the three middle ones are shown from black to
white).}
\label{cubed3}
\end{figure}

The MPI layer is addressed from within \pforest and not exposed to the
\forestclaw code.  The order of leaves is maintained in \pforest according to a
space filling curve.  Each MPI rank has a local view on its own partition,
augmented where necessary with information about one layer of ghost leaves
(see Figure~\ref{cubed3}).

@@

When \forestclaw accesses neighbor patches, they can be on the same or a
different block.  In the latter case, coordinate transformations are carried
out.  The structure of \forestclaw is oblivious to the fact that it only has a
local view of the distributed mesh and data which relieves the developer from
programming to the MPI interface.  Parallel neighbor exchanges are hidden
inside \pforest and called by \forestclaw at well-defined synchronization
points.  There is one such parallel data exchange per time step for a global
value of $\dt$, or one exchange per discretization level per time step if $\dt$
is chosen per-patch depending on its size (this is sometimes called
sub-cycling).  When using sub-cycling, the load balance is attained per level as
it is done for example in Chombo \cite{ColellaGravesKeenEtAl07}
or recent geometric adaptive multigrid schemes
\cite{SundarBirosBursteddeEtAl12}.

The threaded parallelism over the degrees of freedom of a patch can be handled
by \forestclaw alone without the need to involve \pforest.  Additionally many-core implementations, such as \manyclaw \cite{manyclaw}, can be used for the hyperbolic integration on a leaf-patch.  The design of leaf-patches allows for the smart management of data local to accelerator and the host.

@@

For hyperbolic problems, we integrate the solution on a single uniform patch,
containing $m^d$ cells, using the wave propagation algorithm described by R. J.
LeVeque \cite{LeVeque:1997eg} and implemented in \clawpack \cite{le:2002,
clawpack}.  We assume a single degree of freedom per cell and reconstruct a
piecewise constant solution to obtain left and right states at cell edges.  At
each edge, we solve Riemann problems to obtain left and right going waves
propagating at speeds determined from the solution to the Riemann problem.  For
scalar advection, the speed of each wave is simply the local advection speed at
the cell interface.  For non-linear problems and systems, an approximate
Riemann solver, such as a Roe solver \cite{roesolver}, is typically used.  Since
much of the physics of an application can be contained in the Riemann solver,
\forestclaw adopts \clawpack's interface to Riemann solvers effectively allowing
problems solved with \clawpack to be solvable in \forestclaw.  In order to
achieve second order accuracy wave limiters are used to suppress spurious
oscillations at sharp gradients in the flow.

Data exchanges between neighboring patches are done via layers of
ghost cells extending the dimensions along the edges of each patch.
The interior edge values of a given patch overlap the ghost cell
region of a neighboring patch.  For the second order wave propagation
algorithm, two layers of ghost cells are sufficient.  This implies that
one layer of ghost patches is sufficient for $m \ge 4$.  Neighboring
patches at the same level of refinement simply copy their interior
edge values into a neighbors' ghost cells.  Neighboring fine grid
patches average their interior edge data to a coarser neighbor's ghost
cell values.  And neighboring coarse grid patches interpolate data
from their interior edge cells to their fine grid neighbor's ghost
cell values.  To avoid loss of conservation and the creation of
spurious extrema, we use a standard conservative, limited
interpolation scheme to interpolate values from the coarse grid to
fine grid coarse cells \cite{amrclaw, chombo}.  When sub-cycling, time accurate
data between coarse grids is used to fill in ghost cells for fine grids.
As mentioned in the previous section, this procedure can be extended
transparently to distributed parallelism by defining an abstract exchange
routine for ghost patch data.

@@

\section{Numerical results}
Below, we provide two examples that demonstrate the \forestclaw code.
The numerical results to date have been designed to verify that
interface layer between \pforest and \forestclaw is sufficiently
flexible and robust.  Of particular importance was ensuring that all
ghost cell ghost cell transfers (including averaging and
interpolation) have been implemented correctly.
The basic \clawpack algorithm and corresponding code have been
thoroughly tested, and so needed no further verification here.

@@

\noindent
{\bf Example 1 : } In this example, an initial concentration field $q$
is set to 0 in the left half of a computational square and to 1 in the
right half.  A time dependent flow field is prescribed that distorts
the interface between the 0 and 1 concentration values.  The minimum
and maximum levels of refinement were set to 3 and 6, respectively.
This results in a minimum resolution of $64 \times 64$ and a maximum
resolution of $512 \times 512$. \Fig{results2d} shows the results
at an intermediate time step.

% \begin{figure}
% \begin{center}
% \plotbox{\includegraphics[height=0.485\textwidth,clip=true,trim=3cm 0cm 2cm 0cm]{quadtree.png}} \hfil
% \plotbox{\includegraphics[height=0.485\textwidth,clip=true,trim=6cm 0cm 7cm 0cm]{sphere_plot167.png}}
% \end{center}
% \caption{Results for scalar advection using \forestclaw.}
% \label{fig:results}
% \end{figure}

\noindent {\bf Example 2 : } This example, we demonstrate the
multiblock functionality of \forestclaw by considering flow on a
sphere.  The sphere mapping we use consists of two quadtrees, each
defined in the computational space $[-1,1] \times [-1,1]$
\cite{ca-he-le:2008, be-ca-he-le:2009}.  Each quadtree is mapped to cover one hemisphere
of the sphere.  The initial conditions were set to 0 in one half of
the sphere, and 1 in the other half, where the halves are not
necessarily aligned with the equator of the mapping.  The flow field
${u}$ simulates rigid body rotation.  We show the results at an intermediate time in
\Fig{results2d}.

\begin{figure}
\begin{center}
\plotbox{\includegraphics[width=.4\columnwidth]{quadtree}} \hfil
% \hspace{.05\columnwidth}
\plotbox{\includegraphics[width=.4\columnwidth]{sphere_plot167}}
\end{center}
\caption{Numerical results for solving the advection equation.  Left:  Unit
square (single quadtree) with a twisting velocity field.  Right: Spherical ball
with a rotational velocity field constructed from two mapped quadtrees.  In
both cases the concentration is color-coded with a sharp gradient shown in red.
The adaptive mesh refinement follows the location of the gradient (the patches
are not shown where they become too fine for display).  Here we use $m = 8$
cells per \clawpack patch.}
\label{fig:results2d}
\end{figure}

The results in this section are preliminary and will be extended in the final
version.  We have so far exercised \forestclaw for two 2D manifolds, namely the
unit square and a sphere composed of two mapped squares.  On both we solve the
advection equation with a given velocity field.  We advect a concentration
field with a sharp gradient which we track dynamically with adaptive mesh
refinement.  We observe that the refinement follows the front and allows for an
accurate representation of its location over time; see \Fig{results2d}.

\pforest is parallelized with MPI and well tested on large supercomputers, both
in 2D and 3D.  Parallelization of \forestclaw is not completed yet but
envisioned for the final version of the paper.

>>>>

\usepackage{amsmath,amssymb}
\usepackage{cite}
\usepackage{color}
\usepackage{url}
\usepackage{xspace}
\usepackage{graphicx}

@@

% \begin{keyword}
% adaptive mesh refinement,
% wave propagation algorithms
% clawpack,
% HPC,
% manycore
% \end{keyword}

@@

With the advent of high-throughput coprocessors, such as GPGPUs or the MIC
chip, comes the opportunity to sustain unprecedented rates of floating point
operations at comparably high integration density and low cost.  These
architectures, however, require careful structuring of the data layout and
memory access patterns to exhaust their multithreading and vectorization
capabilities.

Consequently, it is not clear a priori how to accelerate PDE solvers
that use adaptive mesh refinement.
%, especially when working with
%unstructured meshes \comment{This sounds we are now going to talk
%  about unstructured meshes.}.
% CB: Good point, took it out.
Of course, it was realized early that it helps to aggregate degrees of
freedom (DOF) at the element level, as has been done with high-order
spectral element \cite{TufoFischer99}
%finite volume (element?)
or discontinuous Galerkin \cite{HesthavenWarburton02}
methods.  GPU implementations of the latter have been proposed recently
\cite{KlocknerWarburtonBridgeEtAl09, BursteddeGhattasGurnisEtAl10}.
The finite volume method has typically been implemented using a single degree
of freedom per cell on
% either
structured
\cite{ppm, clawpack, weno, central schemes}
or unstructured meshes
\cite{ddfv, ausm+, openfoam},
where higher order methods have been constructed by widening the stencil
\cite{needonehere}.
% \comment{Are you looking for
%   references to parallel AMR codes?  Or more general AMR codes based
%   on the FVM?}  \todo{No, I was looking for unstructured FV
%   references.  Would you have any, and can we add the block-AMR
%   references later?} \comment{But most finite volume methods
%   (structured or unstructured) only have a single degree of freedom
%   per mesh cell (``element''?)  Only recently have people increased
%   the order by extending the stencil (without increase the DOFs per
%   mesh cell).}
% CB: Can you fix some of the above citations?

@@

To facilitate hardware acceleration for parallel dynamic AMR, we
build upon the forest-of-octrees paradigm because of its low overhead
and proven scalability \cite{BursteddeWilcoxGhattas11}.  This approach
identifies each octree leaf with a mesh element.  In our work, we go
beyond the traditional high-order element and define each element to
be a dense compute patch with $m^d$ DOFs.  In fact, this approach
resembles \comment{is essentially?} block-structured AMR \cite{be-ol:1984, be-co:1989,
% be-le:1991 (cited below)
ColellaGravesKeenEtAl07,
BerzinsLuitjensMengEtAl10} except that the
patches are not overlapping,
% (except in ghost cell regions),
% CB: we comment on this later in Section 4.
which enables us to capitalize on our
previous experience with scalable FE solvers for PDEs
\cite{BursteddeStadlerAlisicEtAl13}.  The \clawpack software
\cite{LeVeque97} provides a popular implementation of such a patch.
It has been designed to solve hyperbolic conservation laws
% on a uniform compute patch
and successfully used in the context of block-structured AMR
\cite{be-le:1991, amrclaw}.

@@

The starting point of our work is defined by the \pforest algorithms for
forest-of-octrees AMR on the one hand, and the \clawpack algorithms for the
numerical solution of hyperbolic conservation laws
% on uniformly gridded domains
on the other.  Both are specialized codes with the following characteristics:
% \comment{I am not sure about the Python dependency;  the latest version of
% \clawpack relies on
%Python to generate an input file, but the code itself is F77 (or maybe some
%F90). }
% CB: I made it optional.
% DAC: Good.
\begin{center}
\begin{tabular}{l|l|l}
& \multicolumn{1}{c|}{\pforest} & \multicolumn{1}{c}{\clawpack} \\
\hline
subject & hexahedral nonconforming mesh &  hyperbolic PDE on $[0, 1]^d$ \\
toplevel unit & forest of octrees & patch of $m^d$ FV cells \\
atomic unit & octree leaf & one DOF in each cell \\
parallelization & MPI & threads (\manyclaw variant) \\
memory access & distributed & shared on each MPI rank \\
data type & integers & floating point values \\
language & C & Fortran 77 \\
dependencies & none & Python (optional) \\
%\hline
\end{tabular}
\end{center}
Each leaf as the atomic unit of \pforest houses a toplevel unit of \clawpack.
The term cell is used to identify a single DOF within a \clawpack patch.  The
proposed 1:1 correspondence between a leaf and a patch thus combines two
previously disjoint models in a modular way:
\begin{enumerate}
\item We permit the reuse of existing, verified, and performant codes.
% if the integration is done well
\item We preserve the separation between the mesh on one hand and the
discretization and solvers on the other.
\item The AMR metadata (\pforest:
under 1k bytes per octree, 8 bytes per MPI rank,
24 bytes per leaf. \forestclaw: $84 + 28d$ bytes per patch)
% \comment{?? : in 3d: 5 ints,  9 floats; in 2d: 4 ints, 6 floats}
% CB: thanks, I have added it.
is insignificant compared to
the numerical data ($m^d$ floating point values per patch).
\item The resulting parallel programming model is a hybrid (often referred
to as MPI+X).  Only rank-local leaves/patches are stored and computed on.
\end{enumerate}

@@

The MPI layer is addressed from within \pforest and not exposed to the
\forestclaw code.  The order of leaves is maintained in \pforest according to a
space filling curve.  Each MPI rank has a local view on its own partition,
augmented where necessary with information about one layer of ghost leaves.

@@

When \forestclaw accesses neighbor patches, they can be on the same or a
different block.  In the latter case, coordinate transformations are carried
out.  The structure of \forestclaw is oblivious to the fact that it only has a
local view of the distributed mesh and data which relieves the developer from
programming to the MPI interface.  Parallel neighbor exchanges are hidden
inside \pforest and called by \forestclaw at well-defined synchronization
points.  There is one such parallel data exchange per time step for a global
value of $\dt$, or one exchange per discretization level per time step if $\dt$
is chosen per-patch depending on its size (this is sometimes called
subcycling).  When using subcycling, the load balance is attained per level as
it is done for example in Chombo \cite{ColellaGravesKeenEtAl07}
or recent geometric adaptive multigrid schemes
\cite{SundarBirosBursteddeEtAl12}.

The threaded parallelism over the degrees of freedom of a patch can be handled
by \forestclaw alone without the need to involve \pforest.

@@

For hyperbolic problems, we integrate the solution on a a single uniform
patch, containing $m^d$ cells,
using the wave propagation algorithm described by R. J. LeVeque and
implemented in \clawpack \cite{le:2002, clawpack}.  We assume a
single degree of freedom per cell and reconstruct a piecewise
constant solution to obtain left and right states at cell
edges.  At each edge, we solve Riemann problems to obtain left and
right going waves propagating at speeds determined from the solution
to the Riemann problem.  For scalar advection, the speed of each wave
is simply the local advection speed at the cell interface.  For non-linear
problems and systems, an approximate Riemann solver, such as a Roe solver
\cite{roesolver}, is typically used.  Wave limiters are used to surpress
spurious oscillations at sharp gradients in the flow.

Data exchanges between neighboring patches are done via layers of
ghost cells extending the dimensions along the edges of each patch.
The interior edge values of a given patch overlap the ghost cell
region of a neighboring patch.  For the second order wave propagation
algorithm, two layers of ghost cells are sufficient.  This implies that
one layer of ghost patches is sufficient for $m \ge 4$.  Neighboring
patches at the same level of refinement simply copy their interior
edge values into a neighbors' ghost cells.  Neighboring fine grid
patches average their interior edge data to a coarser neighbor's ghost
cell values.  And neighboring coarse grid patches interpolate data
from their interior edge cells to their fine grid neighbor's ghost
cell values.  To avoid loss of conservation and the creation of
spurious extrema, we use a standard conservative, limited
interpolation scheme to interpolate values from the coarse grid to
fine grid coarse cells \cite{amrclaw, chombo}.  When subcycling, time accurate
data between coarse grids is used to fill in ghost cells for fine grids.
As mentioned in the previous section, this procedure can be extended
transparently to distributed parallelism by defining an abstract exchange
routine for ghost patch data.

@@

\section{Numerical results}
Below, we provide two examples that demonstrate the \forestclaw code.
The numerical results to date have been designed to verify that
interface layer between \pforest and \forestclaw is sufficiently
flexible and robust.  Of particular importance was ensuring that all
ghost cell ghost cell transfers (including averaging and
interpolation) have been implemented correctly.  \comment{Parallel performance results will be demonstrated in the full paper?}
The basic \clawpack algorithm and corresponding code have been
thoroughly tested, and so needed no further verification here.

@@

\noindent
{\bf Example 1 : } In this example, an initial concentration field $q$
is set to 0 in the left half of a computational square and to 1 in the
right half.  A time dependent flow field is prescribed that distorts
the interface between the 0 and 1 concentration values.  The minimum
and maximum levels of refinement were set to 3 and 6, respectively.
This results in a minimum resolution of $64 \times 64$ and a maximum
resolution of $512 \times 512$. Figure \Fig{swirl} shows the results
at an intermediate time step.

\begin{figure}
\begin{center}
\plotbox{\includegraphics[height=0.485\textwidth,clip=true,trim=3cm 0cm 2cm 0cm]{quadtree.png}} \hfil
\plotbox{\includegraphics[height=0.485\textwidth,clip=true,trim=6cm 0cm 7cm 0cm]{sphere_plot167.png}}
\end{center}
\caption{Results for scalar advection using \forestclaw.}
\label{fig:results}
\end{figure}

\noindent {\bf Example 2 : } This example, we demonstrate the
multiblock functionality of \forestclaw by considering flow on a
sphere.  The sphere mapping we use consists of two quadtrees, each
defined in the computational space $[-1,1] \times [-1,1]$
\cite{ca-he-le:2008, be-ca-he-le:2009}.  Each quadtree is mapped to cover one hemisphere
of the sphere.  The initial conditions were set to 0 in one half of
the sphere, and 1 in the other half, where the halves are not
necessarily aligned with the equator of the mapping.  The flow field
${u}$ simulates rigid body rotation.  We show the results at an intermediate time in
\Fig{results}.
